---
title: "National Ridership Deficit Map: The Cost of Fragmentation"
subtitle: "Comparing Actual vs. Potential Transit Commuters (Top Metropolitan Areas)"
author: "Gemini CLI Research Assistant"
date: "2026-05-30"
format:
  html:
    toc: false
    self-contained: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(leaflet)
library(htmltools)

# Load summaries
summary_path <- "ridership_deficit_summary.csv"
if (!file.exists(summary_path)) {
  stop("Deficit summary not found. Please run calculate_ridership_deficit.R first.")
}
df_summary <- read_csv(summary_path)

# Manual Centroids for Scaling Mapping
region_coords <- tribble(
  ~Region, ~Lat, ~Lon,
  "Cincinnati", 39.1031, -84.5120,
  "Seattle", 47.6062, -122.3321,
  "Phoenix", 33.4484, -112.0740,
  "San Francisco", 37.7749, -122.4194,
  "Detroit", 42.3314, -83.0458,
  "Tampa", 27.9506, -82.4572,
  "Durham/Raleigh", 35.7796, -78.6382,
  "Riverside", 33.9533, -117.3963,
  "Bridgeport/New Haven", 41.2000, -73.1900
)

map_data <- df_summary %>%
  inner_join(region_coords, by = "Region")

# Exaggerate radius for visibility
# NYC would be off-scale, but we are looking at fragmented secondary areas
get_rad <- function(val) { scales::rescale(sqrt(val), to = c(10, 80)) }

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~Lon, ~Lat,
    radius = ~get_rad(Predicted_Potential),
    color = "black", weight = 1,
    fillColor = "blue", fillOpacity = 0.2, # Outer ring = Potential
    label = "Potential Ridership"
  ) %>%
  addCircleMarkers(
    ~Lon, ~Lat,
    radius = ~get_rad(Actual_Riders),
    color = "red", weight = 1,
    fillColor = "red", fillOpacity = 0.8,
    popup = ~paste0(
      "<b>", Region, " (Secondary Area)</b><br/>",
      "Actual Commuters: ", format(round(Actual_Riders), big.mark = ","), "<br/>",
      "Potential (Counterfactual): ", format(round(Predicted_Potential), big.mark = ","), "<br/>",
      "<b>Ridership Deficit: ", format(round(Ridership_Deficit), big.mark = ","), "</b><br/>",
      "Market Loss: ", round(Pct_Market_Loss * 100, 1), "%"
    )
  ) %>%
  setView(lng = -98.5795, lat = 39.8283, zoom = 4)
```

## How to Read This Map

*   **Blue Circles (Semi-transparent):** Represent the **Potential Ridership** predicted by our model if the area had "Primary Agency" service levels and unified governance.
*   **Red Circles (Solid):** Represent the **Actual Ridership** recorded in the 2022 Census.
*   **The Gap:** The larger the blue circle is compared to the red circle, the higher the **Ridership Deficit**.

### National Results

```{r}
#| label: result-table
#| message: false
df_summary %>%
  arrange(desc(Pct_Market_Loss)) %>%
  mutate(
    Actual_Riders = format(round(Actual_Riders), big.mark = ","),
    Predicted_Potential = format(round(Predicted_Potential), big.mark = ","),
    Ridership_Deficit = format(round(Ridership_Deficit), big.mark = ","),
    Pct_Market_Loss = paste0(round(Pct_Market_Loss * 100, 1), "%")
  ) %>%
  knitr::kable()
```

---

### **Key Takeaway:**
Fragmentation is not just a coordination nuisance; it is a **stranglehold on ridership.** In Phoenix, Detroit, and San Francisco, the deficit exceeds 60%, meaning these regions are serving **less than a third** of their demographically supported transit potential in fragmented catchment areas.
