---
title: "Governance Friction: Quantifying the Impact of Institutional Fragmentation on Transit Ridership"
author: "Benedict Leonardi"
date: today
execute:
  echo: false
  warning: false
---

## Abstract

---

**JEL Codes:** R41, R48, H77, L32  
**Keywords:** Urban Economics, Transit Governance, Institutional Fragmentation, Counterfactual Modeling, Herfindahl-Hirschman Index

---

## I. Introduction

## II. Theoretical Framework: The Coasean Transit Model

$$U = V - C_{fare} - C_{time} - \phi(F)$$

## III. Data and Methodology

### 3.1 Data Sources

### 3.2 The Transit Fragmentation Index (TFI)

$$HHI_{UZA} = \sum_{i=1}^{n} s_i^2$$

$TFI = 1/HHI_{UZA}$

### 3.3 The Counterfactual Deficit Model

## IV. National Fragmentation Trends

```{r}
#| label: results-table
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(kableExtra)

results_path <- "transit_fragmentation_results.csv"

if (file.exists(results_path)) {
  df <- read_csv(results_path)
  
  cities <- c("New York", "Chicago", "Cincinnati", "Houston", "Phoenix", "San Francisco")
  
  df_summary <- df %>%
    filter(map_lgl(`Urbanized Area Name`, ~any(str_detect(.x, cities)))) %>%
    group_by(`Urbanized Area Name`) %>%
    slice_max(Total_UPT, n = 1) %>%
    ungroup() %>%
    arrange(desc(Fragmentation_Index)) %>%
    select(
      `Metropolitan Region` = `Urbanized Area Name`,
      Agencies = Agency_Count,
      `Total Annual Trips` = Total_UPT,
      `TFI` = Fragmentation_Index
    ) %>%
    mutate(
      `Total Annual Trips` = format(as.numeric(`Total Annual Trips`), big.mark = ","),
      `TFI` = format(round(as.numeric(`TFI`), 2), nsmall = 2)
    )

  kable(df_summary, booktabs = TRUE, align = "lccc") %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
}
```

## V. Advanced Analysis: The "Ridership Deficit"

### 5.1 Bias-Corrected Deficit Results

```{r}
#| label: deficit-summary
#| message: false
#| warning: false

deficit_summary_path <- "ridership_deficit_summary.csv"
if (file.exists(deficit_summary_path)) {
  summary_df <- read_csv(deficit_summary_path) %>%
    mutate(
      `Actual Riders` = format(round(Actual_Riders), big.mark = ","),
      `Predicted Potential` = format(round(Predicted_Potential), big.mark = ","),
      `Ridership Deficit` = format(round(Ridership_Deficit), big.mark = ","),
      `Pct. Market Loss` = paste0(round(Pct_Market_Loss * 100, 1), "%")
    ) %>%
    select(Region, `Actual Riders`, `Predicted Potential`, `Ridership Deficit`, `Pct. Market Loss`)
  
  kable(summary_df, booktabs = TRUE, align = "lcccc") %>%
    kable_styling(bootstrap_options = c("striped", "hover"))
}
```

### 5.2 Key Findings: The Fragmentation Tax

## VI. Discussion and Policy Implications

### 6.1 The Cost of Sovereignty

### 6.2 Recommendations for Consolidation

## VII. Conclusion

## VIII. References

*   **Federal Transit Administration (2024).** *2023 National Transit Database.*
*   **U.S. Census Bureau (2022).** *American Community Survey 5-Year Estimates.*
*   **Coase, R. H. (1937).** *The Nature of the Firm.* Economica.
*   **Vuchic, V. R. (2005).** *Urban Transit: Operations, Planning, and Economics.*


---

*[View source on GitHub](https://github.com/bleonardi/Transit_Fragmentation_Research){target="_blank"}*
