---
title: "Mapping Transit Fragmentation Across the US"
subtitle: "Interactive Visualization of Governance Friction (Top 50 Cities)"
author: "Gemini CLI Research Assistant"
date: "2026-05-29"
format:
  html:
    toc: false
    code-fold: true
    self-contained: true
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(leaflet)
library(htmltools)

# 1. Load the Fragmentation Results
results_path <- "transit_fragmentation_results.csv"
if (!file.exists(results_path)) {
  stop("Fragmentation results not found. Please run the analysis script first.")
}
df <- read_csv(results_path)

# 2. Load the Census Gazetteer for Urban Area Centroids
gaz_path <- "data/2020_Gaz_ua_national.txt"
if (file.exists(gaz_path)) {
  df_gaz <- read_tsv(gaz_path) %>%
    select(GEOID, Lat = INTPTLAT, Lon = INTPTLONG) %>%
    mutate(GEOID = str_pad(as.character(GEOID), 5, pad = "0"))
  
  # Join results with coordinates
  map_df <- df %>%
    inner_join(df_gaz, by = c("Urbanized Area Code" = "GEOID"))
} else {
  stop("Gazetteer not found. Ensure 'data/2020_Gaz_ua_national.txt' exists.")
}

# 3. Filter for Top 50 by Ridership
map_df_top50 <- map_df %>%
  arrange(desc(Total_UPT)) %>%
  head(50)

# 4. Create Visualization
# Color palette
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = map_df_top50$Fragmentation_Index
)

# Custom Radius function to make differences visible
# We use a linear scale on the log of trips to exaggerate the difference 
# between Tier 1 (NYC) and Tier 2 (Chicago/SF) vs Tier 3 (Mid-size)
get_radius <- function(upt) {
  # Range of log10(trips) for top 50 is roughly 7.5 to 9.9
  # Map this range to 5 to 40 pixels
  scales::rescale(log10(upt), to = c(5, 40))
}

leaflet(map_df_top50) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~Lon, ~Lat,
    radius = ~get_radius(Total_UPT),
    color = "black", # Outline
    weight = 1,
    fillColor = ~pal(Fragmentation_Index),
    fillOpacity = 0.8,
    popup = ~paste0(
      "<b>", `Urbanized Area Name`, "</b><br/>",
      "Rank by Ridership: ", row_number(desc(Total_UPT)), "<br/>",
      "Agencies: ", Agency_Count, "<br/>",
      "Fragmentation Index: ", round(Fragmentation_Index, 2), "<br/>",
      "Total Annual Trips: ", format(Total_UPT, big.mark = ",")
    )
  ) %>%
  addLegend(
    "bottomright", pal = pal, values = ~Fragmentation_Index,
    title = "Fragmentation Index",
    opacity = 1
  ) %>%
  setView(lng = -98.5795, lat = 39.8283, zoom = 4)
```

## Summary of National Fragmentation (Top 50)

This interactive map focuses on the **Top 50 U.S. Urbanized Areas by ridership**. By limiting the scope, we can more clearly see the relationship between size and governance friction in major metropolitan areas.

*   **Marker Size** is scaled to the logarithm of ridership, with New York City as the maximum reference point.
*   **Marker Color** represents the **Transit Fragmentation Index (TFI)**.
*   **Observations:** Regions with multiple competing heavy-rail or regional rail systems (like the Bay Area) often show higher fragmentation scores than single-provider systems.

### Top 10 Cities in this Dataset

```{r}
#| label: top-table
map_df_top50 %>%
  arrange(desc(Total_UPT)) %>%
  head(10) %>%
  select(`Urbanized Area Name`, Agency_Count, Fragmentation_Index, Total_UPT) %>%
  mutate(Total_UPT = format(Total_UPT, big.mark = ","),
         Fragmentation_Index = round(Fragmentation_Index, 2)) %>%
  knitr::kable()
```
